Stability and accuracy of free surface time integration in buoyancy driven viscous flows

### 2016-05-24 free-surface-paper.tex
Add text on RT test.

Several studies have attempted to alleviate the timestepping requirements imposed by the sloshing 
instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
most free surface calculations have preferred to use explicit timestepping methods for the free 
surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
Stokes matrix, giving it better stability properties.  \citet{popov2008slim3d}, \citet{kramer2012implicit}, and \citet{furuichi2015implicit}
explored methods for solving the transport of the free surface implicitly.

The paper is organized as follows.
After introducing the problem in Section~\ref{sec:governing},
we derive an approach to analyze 
free surface schemes based on the normal modes in Section~\ref{sec:eigenvalue}
and Section~\ref{sec:timestepping}.
In Section~\ref{sec:kmm} we use this approach to look at the
quasi-implicit stabilization proposed in \citet{kaus2010stabilization}.
Then, we propose a new time marching scheme for free surface computations with 
good stability properties (Section~\ref{sec:newscheme}).
Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT} in Section~\ref{sec:implementation}, before showing numerical results
in Section~\ref{sec:results}.

@@

\begin{itemize}
\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. Solution of the whole spectrum is expensive,
however only the minimum relaxation time is required. For certain geometries a simple power iteration on 
the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) can be enough.
\item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosity
variations), an analytical approximation may be sufficient.
\item Scaling. In general, we expect the relaxation times to scale with $\tau \sim \frac{\eta}{\rho g L}$,
where the density, gravity, viscosity, and lengthscale $L$ are all representative values.  A small amount
of experimentation near this value of $\tau$ can find an appropriate relaxation time.
\end{itemize}

@@

In section \label{sec:results} we show examples of the numerical solution of $\tau_\mathrm{min}$
and its evolution in time.

\subsection{Connection with Dynamic Relaxation}
\label{sec:dynamic_relaxation}

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_setup.pdf}
\caption{}
\label{fig:rayleigh_taylor}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor.pdf}
\caption{Rayleigh-Taylor test. (a) Maximum depth of the drip. The blue line is the result using forward Euler (FE) integration of the free surface with a small 500 yr timestep, and is used as a reference solution. The symbols are for several different runs with different timestepping parameters, using both quasi-implicit (QI) and nonstandard finite difference (NSFD) schemes. As the timestepping gets smaller, both schemes converge to the reference solution (though from different directions). (b) Evolution of the minimum relaxation timescale over the course of the simulation. Every 50 timesteps we calculate the minimum relaxation timescale to observe how much it changes during the simulation. It does experience significant changes, but not by orders of magnitude, so an initial estimate is likely sufficient for this case.}
\label{fig:rayleigh_taylor}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_tau_choice.pdf}
\caption{}
\label{fig:rayleigh_taylor}
\end{figure*}

@@

For the uniform refinement cases each point is generated by running the benchmark with a different global refinement level.
For the adaptive case each point is generated by allowing the mesh to be refined, where the maximum refinement level
is limited to the same refinement level of the corresponding global refinement simulation.
In both the uniform and the adaptive cases, the smallest cell size in the most refined case is $\sim$1 km.
We refine every ten timesteps according to gradients in the density and compositional fields.

We have chosen a very small timestep for this test of 500 yr (note that this is significantly smaller than the smallest
relaxation time of $\sim$ 14 kyr), so most timestepping schemes should be stable, including forward Euler.
We choose this step for two reasons: (1) The primary purpose of this test is to investigate the savings
due to using adaptive mesh refinement with a free surface, and (2) It was found in \citet{crameri2012comparison}
that in order for this test to fully converge, a small timestep of several hundred years must be used.

@@

It is not clear that the non-standard finite difference scheme is superior to 
the quasi-implicit scheme. For $\theta=0.5$ the quasi-implicit scheme is more accurate,
but is only conditionally stable. For $\theta=1$ the two schemes are comparably accurate 
in the tests we ran.

>>>>

Several studies have attempted to alleviate the timestepping requirements imposed by the sloshing 
instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
most free surface calculations have preferred to use explicit timestepping methods for the free 
surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
Stokes matrix, giving it better stability properties.  \citet{popov2008slim3d}, \citet{kramer2012implicit}, and \citet{furuichi2015implicit}
explored methods for solving the transport of the free surface implicitly.
 and benchmarks

The paper is organized as follows.
After introducing the problem in Section~\ref{sec:governing},
we derive an approach to analyze 
free surface schemes based on the normal modes in Section~\ref{sec:eigenvalue}
and Section~\ref{sec:timestepping}.
In Section~\ref{sec:kmm} we use this approach to look at the
quasi-implicit stabilization proposed in \citet{kaus2010stabilization}.
Then, we propose a new time marching scheme for free surface computations with 
good stability properties (Section~\ref{sec:newscheme}).
Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT} in Section~\ref{sec:implementation}, 
before showing some numerical results and benchmarks in Section~\ref{sec:results}.

@@

\begin{itemize}
\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. Solution of the whole spectrum is expensive,
however only the minimum relaxation time is required, which means that power iteration on 
the standard eigenvalue problem (Equation~\eqref{eq:standard_eigenvalue}) can be enough.
\item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosity
variations), an analytical approximation may be sufficient.
\item Scaling. In general, we expect the relaxation times to scale with $\tau \sim \frac{\eta}{\rho g L}$,
where the density, gravity, viscosity, and lengthscale $L$ are all representative values.  A small amount
of experimentation near this value of $\tau$ can find an appropriate relaxation time.
\end{itemize}

@@

In section \label{sec:results} we show an examples the numerical solution of $\tau_\mathrm{min}$
and its evolution in time.

\label{sec:dynamic_relaxation}

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_setup.pdf}
\caption{Setup for the Rayleigh-Taylor test. A denser, more viscous layer of lies on top of the mantle. The overlying layer has a thickness of 100 km, and a sinusoidal perturbation on its base with and amplitude of 5 km.}
\label{fig:rayleigh_taylor_setup}
\end{figure*}

We also consider the time evolution of the Rayleigh-Taylor benchmark described in \citet{kaus2010stabilization}.
The benchmark setup is shown in Figure~\ref{fig:rayleigh_taylor_setup}.
In all cases we use adaptive mesh refinement with five global refinement levels and four adaptive refinement levels,
corresponding to $\sim$1 km cells at the most refined level. We refine according to compositional gradients, as well as
near to the free surface.

Figure~\ref{fig:rayleigh_taylor} shows the maximum depth of the interface between the dense and light fluid through time.
The blue line shows the results using forward Euler timestepping with a very small timestep of 500 yrs, and is used as a reference solution.
Nonstandard finite difference timestepping and quasi-implicit timestepping both converge to the reference solution,
though from different directions.
For the nonstandard finite difference simulations we recomputed the value of the minimum relaxation time $\tau_\mathrm{min}$ 
every 50 timesteps using power iteration, and reset the value of $\tau^*$ to $\tau_\mathrm{min}$.

The behavior of the two schemes is similar, though at the same CFL number quasi-implicit timestepping is slightly more accurate.
On the other hand, with our implementation we have found that nonstandard finite difference timestepping can allow for larger timesteps 
than quasi-implicit timestepping while remaining stable.

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor.pdf}
\caption{Rayleigh-Taylor test. (a) Maximum depth of the drip. The blue line is the result using forward Euler (FE) integration of the free surface with a small 500 yr timestep, and is used as a reference solution. The symbols are for several different runs with different timestepping parameters, using both quasi-implicit (QI) and nonstandard finite difference (NSFD) schemes. As the timestepping gets smaller, both schemes converge to the reference solution (though from different directions).}
\label{fig:rayleigh_taylor}
\end{figure*}

We also used the Rayleigh-Taylor benchmark to investigate the effect of making a poor choice for $\tau^*$, shown in Figure~\ref{fig:rayleigh_taylor_tau_choice}.
We include four cases, all using adaptive timestepping with CFL=0.2.
As a reference solution we include the case where we update $\tau^*$ to be equal to $\tau_\mathrm{min}$ every 50 timesteps,
allowing for time variations in the viscosity and density structure to change the minimum relaxation time.
We also show the case where we do not update $\tau^*$ over the course of the simulation, and two cases where
we use the wrong value by a factor of two in either direction.

Overall the course of the Rayleigh-Taylor benchmark, $\tau_\mathrm{min}$ varies by about 25\%, as shown in Figure~\ref{fig:rayleigh_taylor_tau_choice}b.
The case where we use the initial value of $\tau_\mathrm{min}$ for $\tau^*$ gives nearly identical results,
whereas the cases where we choose values $\sim$50\% off in either direction give different (though not wildly so) results.
In this case, using the initial value for $\tau^*$ gives sufficiently accurate results over the course of the simulation.

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/rayleigh_taylor_tau_choice.pdf}
\caption{Effect of the choice of $\tau^*$ on the Rayleigh-Taylor test. (a) Maximum depth of the drip. In all cases adaptive timestepping is used with CFL=0.2. The blue line serves as a reference solution, and shows the result when the smallest relaxation timescale $\tau_{\mathrm{min}}$ is recalculated every 50 timesteps, and then the stabilization timescale $\tau^*$ is set to that value. The red hexagons show the result when we do not update $\tau^*$, instead keeping it at the inital value of $\tau_{\mathrm{min}}$. The results are indistinguishable from those of the first case. The triangles show the results when we make a poor choice of $\tau^*$, by a factor of two in either direction. In these cases there is a significant difference from the reference solution. (b) Evolution of the minimum relaxation timescale over the course of the simulation. It does experience significant changes, but not by orders of magnitude, so it is unsurprising that using the initial value is sufficient for this case.}
\label{fig:rayleigh_taylor_tau_choice}
\end{figure*}

@@

For the uniform refinement cases each point is generated by running the benchmark with a different global refinement level.
For the adaptive case each point is generated by allowing the mesh to be refined according to gradients in density and composition,
where the maximum refinement level is limited to the same refinement level of the corresponding global refinement simulation.
In both the uniform and the adaptive cases, the smallest cell size in the most refined case is $\sim$1 km.
We refine every ten timesteps according to gradients in the density and compositional fields.

We have chosen a very small timestep for this test of 500 yr (note that this is significantly smaller than the smallest
relaxation time of $\sim$ 14 kyr), so most timestepping schemes should be stable and accurate, including forward Euler.
We choose this step for two reasons: (1) The primary purpose of this test is to investigate the savings
due to using adaptive mesh refinement with a free surface, and (2) It was found in \citet{crameri2012comparison}
that in order for this test to fully converge, a small timestep of several hundred years must be used.

@@

It is not clear that the non-standard finite difference scheme is superior to 
the quasi-implicit scheme. For $\theta=0.5$ the quasi-implicit scheme is more accurate,
but has a smaller stability region. For $\theta=1$ the two schemes are comparably accurate,
with a slight advantage to the quasi-implicit scheme. However, we have found that
the nonstandard finite difference scheme allows for larger stable timesteps, if the modeler 
is willing to pay the price of reduced accuracy for a particular simulation.
